---
title: "Electricity demand modelling on GEFCom14 data: solution"
author: "Matteo Fasiolo"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gefcom_small_mgcv}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and have a look at it
```{r loadDat, message = F, fig.width = 8, fig.height = 7}
library(testGam)
data("gefcom_small")

pairs(gefcom_small, lower.panel = NULL)
```


2) Fit a basic Gaussian GAM and visualise the fitted effects
```{r fit_gaus, message = F, fig.width = 8, fig.height = 5}
library(mgcv)
fit1 <- gam(NetDemand ~ NetDemand.24 + Dow + Trend + s(wM) + s(wM_s95) + s(Posan, bs = "cc"), 
             data = gefcom_small)

plot(fit1, all.terms = TRUE, pages = 1)
```
The effect of instantaneous external temperature seems to be the strongest one. 

3) We check the residuals as a function of time (`trend`):
```{r gefsmall2, message = F}
plot(gefcom_small$Trend, residuals(fit1), ylim = c(-1, 1))
```
The long term trend seems to be non-linear. We also check whether we should increase the bases dimension of the other effects:
```{r gefsmall3, message = F, fig.width = 6, fig.height = 6}
par(mfrow = c(2, 2))
gam.check(fit1)
```
`edf` is quite close to `k'` for all effects, and we get a very low p-value for `wM`. We have to increase `k` a bit.

4) Re-fit with a smooth long-term trend and increase basis dimension:
```{r gefsmall4, message = F}
fit2 <- gam(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), 
            data = gefcom_small)

AIC(fit1, fit2)

plot(gefcom_small$Trend, residuals(fit2), ylim = c(-1, 1))
```

AIC is much improved, and there is no systematic residual pattern along `trend`. We recheck using `check`:
```{r gefsmall4a, message = F, fig.width = 6, fig.height = 6}
par(mfrow = c(2, 2))
gam.check(fit2)
```
Maybe we should increase the basis dimension for `wM` a bit more. But increasing the basis dimension for `Trend` too much is definitely not a good idea, because we would end up interpolating the data. 

5) Look at a QQ-plot:
```{r gefsmall5, message = F, fig.width = 4, fig.height = 4}
qq.gam(fit2, rep = 100)
```

We have some fairly fat tails, maybe better to adopt a scaled student-t distribution:
```{r gefsmall6, message = F, fig.width = 4, fig.height = 4}
fit3 <- gam(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), data = gefcom_small, 
             family = scat)

AIC(fit2, fit3)

qq.gam(fit3, rep = 100)
```

Better AIC, and the QQ-plot looks much better. 

6) Worth considering the log-link, as an alternative to the identity link:
```{r gefsmall7, message = F}
fit4 <- gam(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), 
             data = gefcom_small, 
             family = scat(link = log))

AIC(fit3, fit4)
```
Seems better in terms of AIC. We finally plot the fitted effects:
```{r gefsmall8, message = F}
plot(fit4, all.terms = TRUE, pages = 1)
```