---
title: "Big GAM modelling of aggregate GEFCom electricity: solution"
author: "Matteo Fasiolo"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gefcom_big}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and fit a basic Gaussian GAM:
```{r gefbig1, message = F}
library(testGam)
library(mgcViz)
data("gefcom_big")

fit1 <- bamV(NetDemand ~ Dow +
               NetDemand.24 +
               s(Trend, k = 6, bs = "cr") + 
               s(wM, bs = "cr") + 
               s(Instant, bs = "cr") +
               s(wM_s95, bs = 'cr') + 
               s(Posan, bs='cc', k=20),
             data = gefcom_big, 
             aGam = list(discrete = TRUE, nthreads = 2), 
             aViz = list(nsim = 50))

print(plot(fit1, allTerms = TRUE), pages = 1)
```

2. Check if bases are large enough:
```{r bam2, message = F}
tmp <- check(fit1) 
```
The p-values for `Trend` and `Posan` are very low, suggesting that we should increase the dimension of their spline bases (by increasing `k`). However, if we do this for `Trend` (which is simply a time counter) we will ultimately end up interpolating the data! Increasing the basis dimension for `Posan` might lead to a similar problem.

The p-values of `wM`, `wM_s95` and `Instant` are not low, but the Effective Degrees of Freedom (EDF) of the fitted smooths is very close to the bases dimension (`k'`). Hence, we might want to increase $k$ for these effects. Before doing so, we look for residuals patterns along these covariates and `NetDemand.24`:
```{r bam3, message = F}
pl <- check1D(fit1, x = list("NetDemand.24", "wM", "wM_s95", "Instant")) + l_gridCheck1D(n = 40)
print(pl, pages = 1)
```
There is clearly a systematic pattern along `NetDemand.24`, we should clearly use a smooth effect for this variable! Also, `wM` and `wM_s95` show large but irregular deviations,  while along `Instant` we see large deviations during the night (roughly 8am and 6pm).

3. Re-fit with larger bases, re-check and look for interactions
```{r bam4, message = F}
fit2 <- bamV(NetDemand ~ Dow +
             s(NetDemand.24, bs = "cr", k = 20) +
             s(Trend, k = 6, bs = "cr") + 
             s(wM, bs = "cr", k = 20) + 
             s(Instant, bs = "cr", k = 20) +
             s(wM_s95, bs = 'cr', k = 20) + 
             s(Posan, bs='cc', k = 30),
           data = gefcom_big, 
             aGam = list(discrete = TRUE, nthreads = 2), 
             aViz = list(nsim = 50))

tmp <- check(fit2) # Looks better
```
EDF now lower than `k'` for all variables, apart from `Trend` (but this is ok, see above).

```{r bam4b, message = F}
pl <- check1D(fit2, x = list("NetDemand.24", "wM", "wM_s95", "Instant")) + l_gridCheck1D(n = 40)
print(pl, pages = 1)
```

Output of \verb|check1D| looks better than before, especially along `Instant` and `NetDemand.24`.

4) We start by looking for interactions with `Instant`:
```{r bam5, message = F}
pl <- check2D(fit2, 
              x1 = list("NetDemand.24", "wM", "wM_s95", "Posan"), 
              x2 = list("Instant", "Instant", "Instant", "Instant")) + l_gridCheck2D()

print(pl, pages = 1)
```

All of plots show systematic patterns: all interactions with `Instant` need to be included.

```{r bam7, message = F}
fit3 <- bamV(NetDemand ~ Dow +
               s(NetDemand.24, bs = "cr", k = 20) +
               s(Trend, k = 6, bs = "cr") + 
               s(wM, bs = "cr", k = 20) + 
               s(Instant, bs = "cr", k = 20) +
               s(wM_s95, bs = 'cr', k = 20) + 
               s(Posan, bs='cc', k=30) +
               ti(NetDemand.24, Instant, k = c(15, 15), bs = c("cr", "cr")) +
               ti(wM, Instant, k = c(10, 10), bs = c("cr", "cr")) +
               ti(wM_s95, Instant, k = c(15, 15), bs = c("cr", "cr")) +
               ti(Posan, Instant, k = c(10, 15), bs = c("cc", "cr")),
             data = gefcom_big, 
             aGam = list(discrete = TRUE, nthreads = 2), 
             aViz = list(nsim = 50))

AIC(fit2, fit3)
```

Now lets repeat the visual checks
```{r bam8, message = F}
pl <- check2D(fit3, 
              x1 = list("NetDemand.24", "wM", "wM_s95", "Posan"), 
              x2 = list("Instant", "Instant", "Instant", "Instant")) + l_gridCheck2D()

print(pl, pages = 1)
```

Good, the residual patterns are much less visible. 


5. We start by plotting all the marginal effects on one page:
```{r bam10, message = F}
print(plot(fit3, n = 200, select = 1:6), pages = 1)
```

Looks good, but one might want to have a closer look at what happens along `Posan`, which is quite wiggly. Maybe it is possible to do something better to take into account the effect of holidays?  Now we do the same with the 2D interactions:
```{r bam12, message = F}
print(plot(fit3, n2 = 100, select = 7:10) + theme(legend.position="none"), 
      pages = 1)
```

Notice that we used `theme` to remove the legend. We now produce some customized plots:
```{r bam13, message = F}
pl <- plot(fit3, select = 7:10) + 
      l_fitRaster(noiseup = T, pTrans = zto1(0.05, 3, 0.1)) + 
      l_fitContour() + l_points() 

print(pl, pages = 1)
```

Here we used `noiseup = TRUE` to add noise to the fitted surface, which gives an idea about the uncertainty in each effect. We also used `pTrans` to make so that the opacity of the heatmap is proportional to the statistical significance of the effect at each locations. We also set `too.far = -1` to avoid having grey areas far from design points, and we added a sub-sample of only 500 design points using `maxpo = 500`.

The following code produces some RGL plots, here we show only the last one:
```{r bam14, message = F}
plotRGL(sm(fit3, 7), residuals = T)
clear3d()
plotRGL(sm(fit3, 8), residuals = T)
clear3d()
plotRGL(sm(fit3, 9), residuals = T)
clear3d()
plotRGL(sm(fit3, 10), residuals = T)
rglwidget()
```

6. We can look at the variance as follows:
```{r bam15, message = F}
pl <- check2D(fit3, 
              x1 = list("NetDemand.24", "wM", "wM_s95", "Posan"), 
              x2 = list("Instant", "Instant", "Instant", "Instant")) + 
      l_gridCheck2D(gridFun = sd)

print(pl, pages = 1)
```

The variance varies strongly along `Instant`, and it seems to peak around 8am (especially in the winter and at low temperatures). We do the same for skewness:
```{r bam16, message = F}
library(e1071)
pl <- check2D(fit3, 
              x1 = list("NetDemand.24", "wM", "wM_s95", "Posan"), 
              x2 = list("Instant", "Instant", "Instant", "Instant")) + 
      l_gridCheck2D(gridFun = skewness)

print(pl, pages = 1)
```

These plots show that there are some isolated, but large, negative outliers (standardized skewness $\approx -15$). Such drops in demand might be related to public holidays. If we overlook those points, it is quite clear that the residuals are positively skewed, especially in the winter and at low temperatures. We might be able to take this (and the variance) pattern into account using a GAMLSS model (e.g. usign the `gaulss` or `shash` families). But this would take longer to fit, because `bam` is unavailable for such models.
