---
title: "Larynx cancer in Germany: solution"
author: "Matteo Fasiolo, Simon N. Wood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Larynx_mgcv}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and fit a Poisson GAM with MRF smooth:
```{r lary1, message = F, fig.width = 6, fig.height = 6}
library(testGam)
library(mgcv)
data("Larynx")
data("german.polys")

X <- t(sapply(german.polys, colMeans, na.rm=TRUE))

fit1 <- gam(Y ~ s(region, k = 200, bs="mrf", xt=list(polys=german.polys)) + 
                offset(log(E)) + s(x, k=20),
            family=poisson, 
            data=Larynx)

plot(fit1, ask = FALSE)
```

2) Now use an isotropic TPS smooth, and plot the bi-variate effect using two different schemes:
```{r lary2, message = F, fig.width = 8, fig.height = 4}
fit2 <- gam(Y ~ s(X[,1], X[,2], k = 200) + offset(log(E)) + s(x,k=20), 
            family = poisson, data = Larynx)

par(mfrow = c(1, 2))
plot(fit2, select = 1)
plot(fit2, select = 1, scheme = 2)
```

3) Now we use a tensor-product smooth:
```{r lary3, message = F, fig.width = 8, fig.height = 4}
fit3 <- gam(Y~te(X[,1],X[,2],k=c(15, 15))+offset(log(E))+s(x,k=20), 
            family=poisson, data=Larynx)

par(mfrow = c(1, 2))
plot(fit3, select = 1)
plot(fit3, select = 1, scheme = 2)
```

We now compare the three models in terms of AIC
```{r AIC, message = F, fig.width = 8, fig.height = 4}
AIC(fit1, fit2, fit3)
```
The MRF model is better in terms of AIC.

4) This will give an error
```{r bad_practice, message = F, fig.show = 'hide'}
tryCatch(gam.check(fit3), 
         error = function(e) print(e))
```
To solve this we must add the variables use in the model formula to the data.frame
```{r good_practice, message = F, fig.width = 6, fig.height = 6}
Larynx$lon <-  X[,1]
Larynx$lat <-  X[,2]

fit3 <- gam(Y~te(lon, lat,k=c(15, 15))+offset(log(E))+s(x,k=20), 
            family=poisson, data=Larynx)

par(mfrow = c(2, 2))
gam.check(fit3)
```

And now we can also visualise the fit in 3D using
```{r viz_gam, message = F}
vis.gam(fit3, view = c("lon", "lat"), theta = 210, phi = 40 )
```

<!-- 5) Now visualize effect using rgl: -->
<!-- ```{r lary5, message = F} -->
<!-- open3d() -->
<!-- mfrow3d(1, 2) -->
<!-- plotRGL(sm(fit2,1), residuals = TRUE) -->
<!-- next3d() -->
<!-- plotRGL(sm(fit3,1), residuals = TRUE) -->
<!-- rglwidget() -->
<!-- ``` -->
<!-- Tensor effect seems smoother. -->
