---
title: "Mackerel egg data: solution"
author: "Matteo Fasiolo, Simon N. Wood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Mackerel}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
options(rgl.useNULL=TRUE)
options(rgl.printRglwidget=FALSE)
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load packages and plot data
```{r mack1, message = F, fig.width = 6, fig.height = 6}
library(testGam)
library(mgcViz)
data("mack")
data("coast")

## plot data....
with(mack,plot(lon,lat,cex=0.2+egg.dens/150,col="red"))
lines(coast)
ind <- c(1,3,4,5,10,11,16)
pairs(mack[,ind])
```

2) Fit Poisson GAM:
```{r mack2, message = F}
mack$log.net.area <- log(mack$net.area)
fit1 <- gamV(egg.count ~  s(c.dist) + s(I(b.depth^0.5)) + s(temp.surf) + s(temp.20m) + 
                          s(lat) + s(lon) + s(salinity) +
                          offset(log.net.area), data = mack, family = poisson)
```

3) Look at QQ-plot:
```{r mack2a, message = F, fig.width = 4, fig.height = 4}
qq(fit1)
```
does not look good at all, the residuals are over dispersed. Maybe negative binomial is better:
```{r mack3, message = F}
fit2 <- gamV(egg.count ~  s(c.dist) + s(I(b.depth^0.5)) + s(temp.surf) + s(temp.20m) + 
                          s(lat) + s(lon) + s(salinity) +
                          offset(log.net.area), data = mack, family = nb)

qq(fit2)
```

```{r mack4, message = F, fig.width = 4, fig.height = 4}
fit3 <- gamV(egg.count ~  s(c.dist) + s(I(b.depth^0.5)) + s(temp.surf) + s(temp.20m) + 
                          s(lat) + s(lon) + s(salinity) +
                          offset(log.net.area), data = mack, family = tw)

AIC(fit1, fit2, fit3)

qq(fit3)
```
Tweedie seems to be the best.

4) Looking at the residuals as functions of spatial location:
```{r mack5, message = F}
fit3 <- getViz(fit3, nsim = 50)

check2D(fit3, "lon", "lat") + l_gridCheck2D()
```

Refit with a bivariate spatial effect:
```{r mack6, message = F}
fit4 <- gamV(egg.count ~  s(c.dist) + s(I(b.depth^0.5)) + s(temp.surf) + s(temp.20m) + 
                          s(lon, lat, k = 100) + s(salinity) +
                          offset(log.net.area), data = mack, family = tw, 
             aViz = list(nsim = 50))

AIC(fit3, fit4)

check2D(fit4, "lon", "lat") + l_gridCheck2D()
```
Better AIC, and residuals look a bit better across space.

5) Check if bases dimension is adequate:
```{r mack7, message = F}
tmp <- check(fit4)
```
Some p-values are significant, but the `edf` is much lower than `k'`, so maybe no action is required. Further residuals checks don't show anything particularly worrying:
```{r mack8, message = F}
pl <- check1D(fit4, x = list("c.dist", "temp.surf", "temp.20m", "salinity")) + l_gridCheck1D() 
print(pl, pages = 1)
```

6) Plot all smooth effects:
```{r mack9, message = F, fig.width = 9, fig.height = 4}
print(plot(fit4), pages = 1)
```

Look at spatial effect in particular:
```{r mack10, message = F}
plot(sm(fit4, 5)) + l_fitRaster() + l_fitContour() + l_points(colour = 2, shape = 1)
```

Get 3D interactive plot:
```{r mack11a, message = F, echo = 1}
plotRGL(sm(fit4, 5), residuals = TRUE)
rglwidget()
```

<!-- 7) Model with interaction between `b.depth` and spatial location: -->
<!-- ```{r mack11, message = F} -->
<!-- fit5 <- gamV(egg.count ~  s(lon, lat, k = 100) +  -->
<!--                           s(lon, lat, by = I(b.depth^0.5)) + -->
<!--                           s(I(b.depth^0.5)) + -->
<!--                           s(temp.20m) +  -->
<!--                           s(salinity) + -->
<!--                           s(temp.surf) + -->
<!--                           offset(log.net.area), data = mack, family = tw, -->
<!--              aViz = list(nsim = 50)) -->

<!-- AIC(fit4, fit5)  -->
<!-- ``` -->
<!-- Better AIC. Let's look at the spatial effects: -->
<!-- ```{r mack12, message = F} -->
<!-- print(plot(fit5, select = 1:2), pages = 1) -->
<!-- ``` -->