---
title: "Motorcycle accident: solution"
author: "Matteo Fasiolo, Simon N. Wood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{mcycle_mgcv}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and look at it
```{r mcycle, message = F}
library(MASS)
library(mgcv)
data("mcycle")

head(mcycle, 10)

plot(mcycle)
```
Nothing happens until around 15 milliseconds after the impact.

2) We start with a simple Gaussian model
```{r fit_G, message = F}
fitG <- gam(accel ~ s(times), data = mcycle)
```
We now plot the fitted mean, with credible intervals, against the data 
```{r fit_G_pred, message = F}
plot(mcycle)
pr <- predict(fitG, se = TRUE)
lines(mcycle$times, pr$fit)
lines(mcycle$times, pr$fit + 2*pr$se, col = 2)
lines(mcycle$times, pr$fit - 2*pr$se, col = 2)
```
There are several problems with this fit, in particular a) the fit is wiggly where it should be quite flat (`times <  15` and `times >  35`) and b) the intervals have similar width for `times < 15` and for `times > 30` which is clearly inappropriate given that the variance of the residuals increases from left to right. In the following we try to address these issues.

3) Let's look at some diagnostics
```{r gam.check, message = F, fig.show = 'hide'}
par(mfrow = c(2, 2))
gam.check(fitG)
```
We start by considering the text output (the plots produced by `check.gam` have been hidden in this file). We see that `edf` is quite close to `k'`, which suggests that we should increase the number of basis function used. Hence we refit with `k = 20` and re-check
```{r fit_G2, message = F, fig.show = 'hide'}
fitG2 <- gam(accel ~ s(times, k = 20), data = mcycle)
gam.check(fitG2)
```
Now `edf` is well below `k'`, and increasing `k` further seems unnecessary. 

4) We try using an adaptive basis
```{r fit_G2_ad, message = F, fig.show = 'hide'}
fitG2_ad <- gam(accel ~ s(times, k = 20, bs = 'ad'), data = mcycle)
gam.check(fitG2_ad)
```
Notice that `edf` has gone down, relative to the non-adaptive basis. This is because the adaptive basis is able to concentrate all the wiggliness where it is needed (time between 15 and 30 milliseconds), and it produces a nearly linear fit elsewhere. This can be seen by comparing the fitted effects
```{r compare_fits, message = F, fig.width = 8.5, fig.height = 4}
par(mfrow = c(1, 2))
plot(fitG2)
plot(fitG2_ad)
```
Using an adaptive smooth helped us removing some spurious wiggliness for `times > 35` (this was the first issue we mentioned above).

5) Now we start looking at the model residuals
```{r qqgam, message = F}
qq.gam(fitG2_ad, rep = 100)
```
The QQ-plot does not look to bad. However, as we have said before, the data is clearly heteroscedastic. We can see this by looking at the log square residuals
```{r resPlot, message = F}
plot(mcycle$times, log(residuals(fitG2_ad)^2))
```
The variance of the residuals is clearly not constant. 

6) We regress the log squared residuals on `times`,  we plot the effect of time on the log Mean Absolute Deviation (MAD)
```{r fitRes, message = F}
mcycle$res2 <- log( residuals(fitG2_ad, type = "response")^2 )
fitRes <- gam(res2 ~ s(times), data = mcycle)
plot(fitRes)
``` 
As expected the dispersion of the residuals increases with `times`. We obtain the variance by exponentiating the expected value of the log squared residuals and we re-fit with weights inversely proportional to the variance
```{r weight_fit, message = F}
w <- 1 / exp(fitRes$fitted.values) # 1 / variance
fitG2_ad_w <- gam(accel ~ s(times, k = 20, bs = 'ad'), 
                  weights = w,
                  data = mcycle)
``` 
The confidence bands width now nicely changes with times, as can be seen in the following plots
```{r predComp, message = F, fig.width = 8.5, fig.height = 4}
par(mfrow = c(1, 2))
plot(mcycle[ , c("times", "accel")], main = "Unweighted fit")
pr <- predict(fitG2_ad, se = TRUE)
lines(mcycle$times, pr$fit)
lines(mcycle$times, pr$fit + 2*pr$se, col = 2)
lines(mcycle$times, pr$fit - 2*pr$se, col = 2)

plot(mcycle[ , c("times", "accel")], main = "Weighted fit")
pr <- predict(fitG2_ad_w, se = TRUE)
lines(mcycle$times, pr$fit)
lines(mcycle$times, pr$fit + 2*pr$se, col = 2)
lines(mcycle$times, pr$fit - 2*pr$se, col = 2)
``` 


6) We model mean and variance jointly using a Gaussian location scale model
```{r gaulss_fit, message = F, fig.width = 8.5, fig.height = 4}
fit_ls <- gam(list(accel ~ s(times, k=20, bs="ad"), ~s(times)),
              data=mcycle, 
              family=gaulss())

plot(fit_ls, pages=1, scale=0)
``` 


```{r gaulss_pred, message = F}
plot(mcycle[ , c("times", "accel")])
pr <- predict(fit_ls, se = TRUE)
lines(mcycle$times, pr$fit[ , 1])
lines(mcycle$times, pr$fit[ , 1] + 2*pr$se[ , 1], col = 2)
lines(mcycle$times, pr$fit[ , 1] - 2*pr$se[ , 1], col = 2)
``` 