---
title: "Ozone: solution"
author: "Matteo Fasiolo"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Ozone}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and plot it
```{r oz1, message = F}
library(testGam)
library(mgcViz)
data("ozone")
pairs(ozone)
```
2) Fit Gaussian GAM and plot fitted effects:
```{r oz2, message = F}
b <- gamV(O3 ~ s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
              s(dpg)+s(ibt)+s(vis)+s(doy),family=gaussian(link=log),
              data=ozone)
print(plot(b), pages = 1)
```

3) Check model residuals:
```{r oz3, message = F}
qq(b)
check(b)
```
The variance increases with the mean.

4) Refit using the Gamma response distribution:
```{r oz4, message = F}
b <- gamV(O3~s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
            s(dpg)+s(ibt)+s(vis)+s(doy),family=Gamma(link=log),
            data=ozone)

check(b)
```
Looks better!

5) Comparing log link with identity link:
```{r oz5, message = F}
b1 <- gamV(O3~s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
            s(dpg)+s(ibt)+s(vis)+s(doy),family=Gamma(link=identity),
            data=ozone)

AIC(b,b1) ## 
```
multiplicative (log link) was better (lower AIC).

6) Plotting all effects:
```{r oz6, message = F}
print(plot(b), pages=1)
summary(b)
```
Drop ibd, which has the highest p-value:
```{r oz7, message = F}
b1 <- gamV(O3~s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
              s(dpg)+s(vis)+s(doy),family=Gamma(link=log),
              data=ozone)
summary(b1)
```
Try to drop humidity too:
```{r oz8, message = F}
b2 <- gamV(O3~s(vh)+s(wind)+s(temp)+s(ibh)+
              s(dpg)+s(vis)+s(doy),family=Gamma(link=log),
              data=ozone)
summary(b2)
```

Comparing AICs:
```{r oz9, message = F}
AIC(b1, b2)
```
b1 has marginally higher AIC, but choose b2 for simplicity.

7) Plot effects of final model:
```{r oz10, message = F}
print(plot(b2), pages = 1)
```

