---
title: "Ozone: solution"
author: "Matteo Fasiolo, Simon N. Wood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Ozone_mgcv}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and plot it
```{r oz1, message = F, fig.width = 8, fig.height = 7}
library(testGam)
library(mgcv)
data("ozone")
pairs(ozone)
```
2) Fit Gaussian GAM and plot fitted effects:
```{r oz2, message = F, fig.width = 8, fig.height = 8}
b <- gam(O3 ~ s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
              s(dpg)+s(ibt)+s(vis)+s(doy),family=gaussian(link=log),
              data=ozone)
plot(b, pages = 1)
```

3) Check model residuals:
```{r oz3, message = F, fig.width = 8, fig.height = 6}
par(mfrow = c(2, 2))
gam.check(b)
```
The residuals vs linear predictor plot clearly shows that the variance increases with the mean.

4) Refit using the Gamma response distribution:
```{r oz4, message = F, fig.width = 8, fig.height = 6}
b <- gam(O3~s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
            s(dpg)+s(ibt)+s(vis)+s(doy),family=Gamma(link=log),
            data=ozone)

par(mfrow = c(2, 2))
gam.check(b)
```
The residuals vs linear predictor and the QQ-plot look better. In the response vs fitted values plot we still see that the variance of the response increases with the fitted values (the mean), but this is being modelled via the gamma distribution (as demonstrated by the residuals vs linear predictor plot).

5) Comparing log link with identity link:
```{r oz5, message = F}
b1 <- gam(O3~s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
            s(dpg)+s(ibt)+s(vis)+s(doy),family=Gamma(link=identity),
            data=ozone)

AIC(b,b1) ## 
```
The multiplicative model (log link) was better (lower AIC).

6) Plotting all effects:
```{r oz6, message = F, fig.width = 8, fig.height = 8}
plot(b, pages=1)
summary(b)
```
Drop `ibd`, which has the highest p-value:
```{r oz7, message = F}
b1 <- gam(O3~s(vh)+s(wind)+s(humidity)+s(temp)+s(ibh)+
              s(dpg)+s(vis)+s(doy),family=Gamma(link=log),
              data=ozone)
summary(b1)
```
Try to drop humidity too:
```{r oz8, message = F}
b2 <- gam(O3~s(vh)+s(wind)+s(temp)+s(ibh)+
              s(dpg)+s(vis)+s(doy),family=Gamma(link=log),
              data=ozone)
summary(b2)
```

Comparing AICs:
```{r oz9, message = F}
AIC(b1, b2)
```
b1 has marginally higher AIC, but choose b2 for simplicity.

7) Plot effects of final model:
```{r oz10, message = F, fig.width = 8, fig.height = 8}
plot(b2, pages = 1)
```

