---
title: "Rent modelling in Munich: solution"
author: "Matteo Fasiolo"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Rent_munich_GAMLSS}
  %\VignetteEncoding{UTF-8}
---

  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data, plot it and fit a basic GAM:
```{r dbbmi1, message = F, fig.width = 8, fig.height = 7}
library(testGam)
library(mgcViz)
data("munich_rent")

pairs(munich_rent, lower.panel = NULL)

fit1 <- gamV(R ~ s(Fl) + s(A) + loc + B + H + L, data = munich_rent, aViz = list(nsim = 50))

summary(fit1)
```

Plot smooth effects:
```{r dbbmi2, message = F, fig.width = 8, fig.height = 4}
print(plot(fit1), pages = 1)
```

2) Effect of `Fl` looks linear, and an interaction with location makes sense:
```{r dbbmi2a, message = F}
fit2 <- gamV(R ~ Fl + Fl:loc + s(A) + loc + B + H + L, data = munich_rent, aViz = list(nsim = 50))

AIC(fit1, fit2)
```

We get lower AIC. See if coefficients make sense:
```{r dbbmi3, message = F}
summary(fit2)
```
Effect of floor space `Fl` must increase with the location desirability `loc`.

3) Looking at variance of residuals across covariates:
```{r dbbmi4, message = F, fig.width = 8, fig.height = 5}
pl <- check1D(fit2, list("Fl", "A", "B", "H", "L", "loc")) + l_gridCheck1D(gridFun = sd)
print(pl, pages = 1)
```
The variance depends on pretty much all covariates, with possibly the exception of `A` and `loc`. Fit a Gaussian GAMLSS model:
```{r dbbmi5, message = F}
fit3 <- gamV(list(R ~ Fl + Fl:loc + s(A) + loc + B + H + L, 
                    ~ Fl + Fl:loc + s(A) + loc + B + H + L), 
             data = munich_rent, family = gaulss, aViz = list(nsim = 50))

AIC(fit2, fit3)
```
Seems an improvements. Re-check residuals variance:
```{r dbbmi6, message = F}
pl <- check1D(fit3, list("Fl", "A", "B", "H", "L", "loc")) + l_gridCheck1D(gridFun = sd)
print(pl, pages = 1)
```
Looks better. Let's look at the significance of the effects:
```{r dbbmi7, message = F}
summary(fit3)
```
Marginal effect of `loc` is not significant in the mean model, but we leave it there because the interaction `Fl:loc3` seems important.

4) Looking at residuals distribution:
```{r dbbmi8, message = F}
qq(fit3)
```
It is skewed to the right. Let's look at skewness as a function of the covariates:
```{r dbbmi9, message = F}
library(e1071)
pl <- check1D(fit3, list("Fl", "A", "B", "H", "L", "loc")) + l_gridCheck1D(gridFun = skewness)
print(pl, pages = 1)
```

We adopt a `shash` models and include a model for skewness:
```{r dbbmi10, message = F}
library(mgcFam)
fit4 <- gamV(list(R ~ Fl + Fl:loc + s(A) + loc + B + H + L, 
                    ~ Fl + Fl:loc + s(A) + loc + B + H + L, 
                    ~ Fl + Fl:loc + s(A) + loc + B + H + L, 
                    ~ 1), 
             data = munich_rent, family = shash, aViz = list(nsim = 50))

AIC(fit3, fit4)
```
Seems to improve things in terms of AIC. Lets re-check the skewness:
```{r dbbmi11, message = F}
pl <- check1D(fit4, list("Fl", "A", "B", "H", "L", "loc")) + l_gridCheck1D(gridFun = skewness)
print(pl, pages = 1)
```
Looks better (notice that `l_gridCheck1D` by default uses 80\% intervals). Final check:
```{r dbbmi12, message = F}
check(fit4)
```

Residuals distribution looks good, maybe we should increase `k` for the year of construction effect `A`:
```{r dbbmi13, message = F}
print(plot(fit4), pages = 1)
```
Increasing `k` does not seem reasonable, especially for skewness where this effect is not significant:
```{r dbbmi14, message = F}
summary(fit4)
```