---
title: "Retinopathy among diabetics: solution"
author: "Matteo Fasiolo, Simon N. Wood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Retinopath}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and look at it:
```{r ret1, message = F}
library(testGam)
library(mgcViz)
data("wesdr")

pairs(wesdr)
```

2) Fit logistic model with all interactions
```{r ret2, message = F}
k = 7
fit1 <- gamV(ret ~ ti(dur,k=k) + ti(gly,k=k) + ti(bmi,k=k) + ti(dur,gly,k=k)+
                   ti(dur,bmi,k=k) + ti(gly,bmi,k=k), 
             data=wesdr,
             family=binomial(), 
             aGam = list(select=TRUE),
             aViz = list(nsim = 50))

summary(fit1)
```
Only last interaction seems important. Plot marginal effects:
```{r ret3, message = F}
print(plot(fit1, select = 1:3), pages = 1)
```

Plot interaction effects:
```{r ret4, message = F}
print(plot(fit1, select = 4:6), pages = 1)
```
First two interactions are essentially zero.

3) Fit simplified model and do residuals checks:
```{r ret5, message = F}
fit2 <- gamV(ret ~ s(dur,k=10) + te(gly,bmi,k=c(10, 10)), 
             data=wesdr,
             family=binomial(),
             aViz = list(nsim = 50))

pl <- check1D(fit2, list("dur", "gly", "bmi")) + l_gridCheck1D(level = 0.95)
print(pl, pages = 1)
```
Looks relatively good. Checking across two variables:
```{r ret6, message = F}
check2D(fit2, "gly", "bmi") + l_gridCheck2D()
```
Big residuals in two locations... not sure it is worth increasing k.

QQ-plot with normally transformed residuals:
```{r ret7, message = F}
qq(fit1, type = "tnorm")

# shine(qq(fit1, type = "tnorm")) # cannot do it in the html, but do try this
```

4) Plot tensor effect interactively:
```{r ret8, message = F}
plotRGL(sm(fit2, 2), residuals = TRUE)
rglwidget()
```
