---
title: "Retinopathy among diabetics part 2: solution"
author: "Matteo Fasiolo, Simon N. Wood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Retynopathy_mgcv_2}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and look at it:
```{r ret1, message = F}
library(testGam)
library(mgcv)
data("wesdr")

pairs(wesdr)
```

2) Fit logistic model with all interactions
```{r ret2, message = F}
k = 10
fit1 <- gam(ret ~ ti(dur,k=k) + ti(gly,k=k) + ti(bmi,k=k) + ti(dur,gly,k=k)+
                  ti(dur,bmi,k=k) + ti(gly,bmi,k=k), 
            data=wesdr,
            family=binomial())

summary(fit1)
```
None of the interactions seem very important. We now plot all effects
```{r ret3, message = F, fig.width = 8, fig.height = 6}
plot(fit1, pages = 1)
```
The interaction between `dur` and `bmi` seems to be overfitting quite horribly, and the marginal effect of of `gly` is exploding for `gly > 17`. This does not look good. 

3) Refit using REML for smoothing parameter selection
```{r REML, message = F}
fit2 <- gam(ret ~ ti(dur,k=k) + ti(gly,k=k) + ti(bmi,k=k) + ti(dur,gly,k=k)+
                  ti(dur,bmi,k=k) + ti(gly,bmi,k=k), 
            data=wesdr,
            family=binomial(), method = "REML")

summary(fit2)
```
The number of EDF used for all effects has decreased greatly. And the effects look much more reasonable
```{r plotREML, message = F, fig.width = 8, fig.height = 6}
plot(fit2, pages = 1)
```


4) Now we try to select the effects by penalising also their parametric part
```{r ret4, message = F}
fit3 <- gam(ret ~ ti(dur,k=k) + ti(gly,k=k) + ti(bmi,k=k) + ti(dur,gly,k=k)+
                  ti(dur,bmi,k=k) + ti(gly,bmi,k=k), 
            data=wesdr,
            family=binomial(), 
            select = TRUE, method = "REML")

summary(fit3)
```
Notice that the effective degrees of freedom of the first two interaction are close to zero while, without `select = TRUE`, they could not go below 1 (the linear component of the interaction, which is unpenalised). In fact, we see that the second interaction is now almost flat
```{r plotInt, message = F}
plot(fit3, select = 5)
```

5) Fit simplified model and look at the effects:
```{r retSimple, message = F, fig.width = 7, fig.height = 7}
fit4 <- gam(ret ~ s(dur, k=10) + ti(gly, k=10) + ti(bmi, k=10) + ti(gly, bmi,k = 10), 
            data=wesdr,
            family=binomial(), method = "REML")

plot(fit4, pages = 1)
```

Compare with model with simple linear interaction
```{r retLinear, message = F}
fit5 <- gam(ret ~ s(dur, k=10) + ti(gly, k=10) + ti(bmi, k=10) + I(gly * bmi), 
            data=wesdr,
            family=binomial(), method = "REML")

AIC(fit4, fit5)
```
The tensor interaction seems slightly better.



<!-- 4) Plot tensor effect interactively: -->
<!-- ```{r ret8, message = F} -->
<!-- #plotRGL(sm(fit2, 2), residuals = TRUE) -->
<!-- rglwidget() -->
<!-- ``` -->
